Too many colors spoil the graph

### 2023-10-10 text/20/poor-color-choice.Rmd

![](http://www.pmean.com/images/20/misuse-of-gradient.png)

@@

```{r color-circle1, echo=FALSE, fig.width=5, fig.height=5}
suppressMessages(suppressWarnings(library(ggplot2)))
suppressMessages(suppressWarnings(library(RColorBrewer)))
n <- 500
k <- 5
delta <- 0.9
angle_seq <- (1:n)/n
df0<- data.frame(x=delta*sin(2*pi*angle_seq), y=delta*cos(2*pi*angle_seq))
sb <- (n/k)*(1:k)
df1 <- df0[sb, ]

ggplot(df0, aes(x, y)) +
  geom_point(color=hsv(angle_seq, 1, 1), size=2, shape=16) +
  geom_point(data=df1, aes(x, y), color=hsv(sb/n, 1, 1), size=24, shape=16) +
  expand_limits(x=c(-1, 1), y=c(-1, 1)) +
  theme_void() +
  theme(
    panel.background=element_rect(fill="#808080"))
```

@@

```{r color-circle2, echo=FALSE, fig.width=5, fig.height=5}
ggplot(df0, aes(x, y)) +
  geom_point(color=hsv(angle_seq, 1, 0.5), size=2, shape=16) +
  geom_point(data=df1, aes(x, y), color=hsv(sb/n, 1, 0.5), size=24, shape=16) +
  expand_limits(x=c(-1, 1), y=c(-1, 1)) +
  theme_void() +
  theme(
    panel.background=element_rect(fill="#808080"))
```

```{r color-circle3, echo=FALSE, fig.width=5, fig.height=5}
ggplot(df0, aes(x, y)) +
  geom_point(color=hsv(angle_seq, 0.5, 1), size=2, shape=16) +
  geom_point(data=df1, aes(x, y), color=hsv(sb/n, 0.5, 1), size=24, shape=16) +
  expand_limits(x=c(-1, 1), y=c(-1, 1)) +
  theme_void() +
  theme(
    panel.background=element_rect(fill="#808080"))
```

@@

```{r brew1, fig.height=1}
par(mar=rep(0, 4))
display.brewer.pal(5, "Set1")
brewer.pal(5, "Set1")
```

These aren't taken directly from a color wheel but they are well spaced out from a visual perspective, meaning that it is easy to distinguish among these different colors.

>>>>

![](http://www.pmean.com/new-images/20/poor-color-choice-01.png)

@@

```{r color-circle1, echo=FALSE, fig.width=5, fig.height=5}
suppressMessages(suppressWarnings(library(ggplot2)))
suppressMessages(suppressWarnings(library(RColorBrewer)))
n <- 500
k <- 5
delta <- 0.9
angle_seq <- (1:n)/n
df0<- data.frame(
    x=delta*sin(2*pi*angle_seq),
    y=delta*cos(2*pi*angle_seq))
sb <- (n/k)*(1:k)
df1 <- df0[sb, ]

ggplot(df0, aes(x, y)) +
  geom_point(
      color=hsv(angle_seq, 1, 1), 
      size=2, 
      shape=16) +
  geom_point(
      data=df1, 
      aes(x, y), 
      color=hsv(sb/n, 1, 1), 
      size=24, 
      shape=16) +
  expand_limits(
      x=c(-1, 1), 
      y=c(-1, 1)) +
  theme_void() +
  theme(
      panel.background =
          element_rect(fill="#808080"))
```

@@

```{r color-circle2, echo=FALSE, fig.width=5, fig.height=5}
ggplot(df0, aes(x, y)) +
  geom_point(
      color=hsv(angle_seq, 1, 0.5), 
      size=2, 
      shape=16) +
  geom_point(
      data=df1, 
      aes(x, y), 
      color=hsv(sb/n, 1, 0.5), 
      size=24, 
      shape=16) +
  expand_limits(
      x=c(-1, 1), 
      y=c(-1, 1)) +
  theme_void() +
  theme(
      panel.background = 
          element_rect(fill="#808080"))
```

You might also consider lighter colors. These are pastel colors. Going lighter or going darker will depend on many factors, but either choice is usually better than using the pure colors.

```{r color-circle3, echo=FALSE, fig.width=5, fig.height=5}
ggplot(df0, aes(x, y)) +
  geom_point(
      color=hsv(angle_seq, 0.5, 1), 
      size=2, 
      shape=16) +
  geom_point(
      data=df1, 
      aes(x, y), 
      color=hsv(sb/n, 0.5, 1), 
      size=24, 
      shape=16) +
  expand_limits(
      x=c(-1, 1), 
      y=c(-1, 1)) +
  theme_void() +
  theme(
      panel.background =
          element_rect(fill="#808080"))
```

@@

```{r brew1, fig.height=1}
par(mar=rep(0, 4))
display.brewer.pal(5, "Set1")
pal5 <- brewer.pal(5, "Set1")
```

These aren't taken directly from a color wheel but they are well spaced out from a visual perspective, meaning that it is easy to distinguish between any two pairs of these different colors. You can find similar choices in Python, SAS and Stata.

SPSS does not have specific color palettes for nominal data, but it does organize its colors so that most of the rows are good choices for representing nominal data and most of the columns are good choices for representing ordinal data.

![](http://www.pmean.com/new-images/20/poor-color-choice-02.png)

Ordinal data is often well represented by a gradient of dark to light levels of a particular color, but this is a poor choice for nominal data. With nominal data, you want a good amount of separation between any two colors. You can get this by selecting evenly spaced values on a color wheel.

### 2023-10-10 text/23/too-many-colors.rmd

When you draw a graph, either a bar chart, a scatterplot, or even a pie chart, you have the choice of a broad range of colors that you can use. R, for example, has 657 different colors from aliceblue to yellowgreen. SAS has 13 shades of orange, 33 shades of blue, and 47 shades of green. They even have different shades of black. You have a wealth of colors, but you can't sue all of them in the same graph. The ideal number of colors is 2.

@@

Here's another lesson that I learned from Naomi Robbins, another expert on data visualization. She points out that you don't make every word in a sentence a different color. You use a single color to draw emphasis to a particular word or phrase. The same should be true for your graphs.

@@

```{r, echo=FALSE, warning=FALSE}
us +
  scale_fill_continuous(
        low = "green", 
        high = "blue", 
        name = "Poverty rate (2014)", 
        label = scales::comma) + 
  theme(legend.position = "right")
```

@@

First, the ends of the rainbow do not show enough contrast. In the above map, the low poverty rate of New Hampshire (9.2%) is a red that differs only slightly from high poverty rate of Mississippi (21.9%) shown as magenta. If anything, the contrast is bigger between Mississippi (magenta) and Alabama (blue) even though the poverty rates of 19.2% for Alabama is a lot closer than New Hampshire's.

>>>>

When you draw a graph, either a bar chart, a scatterplot, or even a pie chart, you have the choice of a broad range of colors that you can use. R, for example, has 657 different colors from aliceblue to yellowgreen. SAS has 13 shades of orange, 33 shades of blue, and 47 shades of green. They even have different shades of black. You have a wealth of colors, but you can't use all of them in the same graph. The ideal number of colors is 2. Sometimes 3 is okay, but usually 2 is better.

@@

Here's a lesson that I learned from Naomi Robbins, another expert on data visualization. She points out that you don't make every word in a sentence a different color. You use a single color to draw emphasis to a particular word or phrase. The same should be true for your graphs.

@@

@@

First, the ends of the rainbow do not show enough contrast. In the above map, the low poverty rate of New Hampshire (9.2%) is a red that differs only slightly from high poverty rate of Mississippi (21.9%) shown as magenta. If anything, the contrast is bigger between Mississippi (magenta) and Alabama (blue) even though the poverty rate of 19.2% for Alabama is a lot closer than New Hampshire's.
